```{r}
#install.packages("dplyr")
library(dplyr)
library(ggplot2)
```

We will be practicing with `dplyr` functions using the `starwars` dataset:

```{r}
head(starwars)
```

### Subset rows with `filter`:

`filter()` allows us to subset **rows** of a dataset based on the values.

```{r}
## Filter for rows with human Star Wars characters.
filter(starwars, species == "Human")

## save to a new R object
humans <- filter(starwars, species == "Human")
male <- filter(humans, sex == "Male")

## filter mutiple columns at once
filter(starwars, species == "Human", sex == "Male")
```

We can also use different comparison operators to filter:

-   `!` is "not"

-   `==` is "equal"

-   `&`is "and"

-   `|` is "or"

-   `<`is "less than"

-   `>`is "greater than"

-   `is.na()` returns TRUE if value is NA

```{r}
## return characters with missing mass
filter(starwars, is.na(mass))
```

```{r}
## how to get all characters that are NOT human?
filter(starwars, species != "Human")
```

variable_name `%in%` vector_of_values returns TRUE if the value of the variable name is equal to one of the values in vector_of_values

```{r}
"yellow"%in%c("yellow", "fair", "tan")

##how would we filter for characters with brown or blond hair
filter(starwars, hair_color %in% c("blond", "brown"))
```

`grepl("characters_of_interest", variable_name)` returns TRUE if the value of `variable_name` contains the `character_of_interest`

```{r}
filter(starwars, grepl("blond", hair_color))
#how would we get all characters with "blue" in their eyes
filter(starwars, grepl("blue", eye_color))
```

You can use `<,` `>,` `<=,` or `>=` to compare numerical or categorical variables.

-   Nominal categorical variables (character vectors) are ranked alphabetically

-   Ordinal categorical variables (factors) are ranked based on the built in order

```{r}
## starwars characters with  a name starting with letters before "D"
filter(starwars, name < "D") #ranking is based on alphabetical order

str(diamonds)
filter(diamonds, cut > "Good") #ranking is based on the levels of the ordered factor cut

"A" > "z" #FALSE

```

between(x, left, right) is a shorthand for x\>=left & x\<=right

```{r}
between(5,1,7) #TRUE
filter(starwars, between(mass, 50, 100))
```

```{r}
filter(starwars, is.na(hair_color))

filter(starwars, height < 100, species != "Droid")

filter(starwars, between(birth_year, 100,200))

filter(starwars, grepl("red", hair_color))
```

### Order rows with `arrange`

`arrange() reorders rows based on the values of column(s)`

```{r}
##order by character name
arrange(starwars, name)
#order of chracter values aA, bB, cC

##order by species and birth year
arrange(starwars, species, birth_year)

## tallest character
arrange(starwars, desc(height))

## shortest chracter
arrange(starwars, height, species)

#
arrange(starwars, hair_color, eye_color)
```

### Subset columns with `select()`

```{r}
character_colors <- select(starwars, hair_color, skin_color, eye_color)
```

```{r}
#all columns + name and eye_color inclusive
select(starwars, name : eye_color)
```

We can also remove columns using a `-` sign.

```{r}
select(starwars, -films, - vehicles, -starships)
select(starwars, -c(films, vehicles, starships))
select(starwars, -c(films:starships))
```

**More helpful functions**

`everything()` is a shortcut for "everything else"

```{r}
select(starwars, species, everything())

##alphabetize columns
starwars_names <- sort(names(starwars))
select(starwars, starwars_names)
  
```

`starts_with("abc")` matches column names that start with "abc"

`ends_with("abc")` matches column names that end with "abc"

`contains("abc")` matches column names that contain "abc"

```{r}
select(starwars, starts_with("h"))
select(starwars, contains("or"))
```

Practice!

```{r}
#select each character's homeworld and species
select(starwars, homeworld, species)
# Could be useful for what world can typically be attributed to a certain species.

select(starwars, hair_color:birth_year)
```

### Add new variables with `mutate()`

```{r}
starwars_small <- select(starwars, height, mass, species)

mutate(starwars_small, 
       height_m = height/100)
```

We can also create multiple variables at once and we can refer to variables we just created.

```{r}
mutate(starwars_small,
       height_m = height/100,
       bmi = mass/height_m^2)

```

Group a variable using the `.by` parameter

```{r}
mutate(starwars_small,
       species_mass = mean(mass, na.rm=T),
       .by = species)

```

Practice!

```{r}
mutate(starwars_small,
       weight_lbs = mass*2.2)
```
